---
output:
  word_document: default
  html_document: default
---
Homework 6: Question 6
>Q6. How would you generalize the original code below to work with any set of input protein structures?

hc <- hclust( dist( rbind(s1.b, s2.b, s3.b) ) )
plot(hc)

Updated Code: 
```{r}
analyze_and_cluster_bfactors <- function(pdb_files, chain = "A", elety = "CA") {
  # Load, trim, and extract B-factors for all input PDB files
  b_factors <- lapply(pdb_files, function(file) {
    trim.pdb(read.pdb(file), chain = chain, elety = elety)$atom$b
  })
  
  # Perform hierarchical clustering and plot the dendrogram
  hc <- hclust(dist(do.call(rbind, b_factors)))
  plot(hc, main = "B-factor Clustering", xlab = "Proteins", sub = "")
}
```

# Example: Provide a set of PDB file names from previous questons in section 2

```{r}
analyze_and_cluster_bfactors(c("4AKE", "1AKE", "1E4Y"))
```

##Documentation Rubric:
>Inputs of the functions:
  -pdb_files: Character vector containing the name of the PBD files 
  -chain: identifies chain to extract (default = "A").
  -elety: atom type to keep (default = "CA").

>What does the function do?:
- Reads many PBD files to then extract the B-factors from the specific chain and atom type. The function then clusters the data to compare.  
  
>Output:
Plot shows clustering of the input proteins based on similiarty in B-factors. 

##Code Rubric: 
>Correct Functionality 

