---
title: "class 10: halloween"
author: "Mia Fava"
format: pdf
---

```{r}
library(knitr)
```



##Importing Candy Data
```{r}
candy_file <- "class10.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```

>Q1. How many different candy types are in this dataset?

```{r}
num_candies <- nrow(candy)
print(num_candies)
```
85 Different types

>Q2. How many fruity candy types are in the dataset?

```{r}
table(candy$fruity)
num_fruity <- sum(candy$fruity == 1)
```
There are 38 fruity candies 

##What's your favorite candy?

```{r}
candy["Twix", ]$winpercent
```

>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
favorite_candy <- "Swedish Fish"
favorite_winpercent <- candy[favorite_candy, "winpercent"]
print(favorite_winpercent)
```
My favorite candy is Swedish Fish with a winpercent of 54.9%

>Q4. What is the winpercent value for “Kit Kat”?

```{r}
kitkat_winpercent <- candy["Kit Kat", "winpercent"]
print(kitkat_winpercent)
```

Winpercent of Kit Kat is 76.8%

>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
tootsie_winpercent <- candy["Tootsie Roll Snack Bars", "winpercent"]
print(tootsie_winpercent)
```

Winpercent of Toosie Roll Snack Bars is 49.7%


>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

```{r}
library(skimr)
skimr::skim(candy)
```

The column that seems to be on a different scale is likely the winpercent due to how it is measured as a percentage. The other columns are measured as binary values, while winpercent has a range of values that are seen to be continuous. 

>Q7. What do you think a zero and one represent for the candy$chocolate column?

```{r}
unique(candy$chocolate)
```

For this column, the 0 infers that the candy does not contain chocolate and a 1 means that the candy contains chocolate. 

>Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent, 
     main = "Histogram of Winpercent Values", 
     xlab = "Winpercent", 
     col = "red", 
     border = "black")
```

>Q9. Is the distribution of winpercent values symmetrical?

The histrgram shows that the distribution of winpercent are not perfectly symmetrical, there is a slight skew to the left as seen in the graph above.  

>Q10. Is the center of the distribution above or below 50%?

```{r}
median_winpercent <- median(candy$winpercent)
print(median_winpercent)
```
 
The center is the at 48%. 

>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
chocolate_winpercent <- mean(candy$winpercent[as.logical(candy$chocolate)])
print(chocolate_winpercent)
```

```{r}
fruity_winpercent <- mean(candy$winpercent[as.logical(candy$fruity)])
print(fruity_winpercent)
```

Chocolate is ranked higher as 60.92% > 44.12%. 

>Q12. Is this difference statistically significant?

```{r}
t_test_result <- t.test(candy$winpercent[as.logical(candy$chocolate)],                         candy$winpercent[as.logical(candy$fruity)])
print(t_test_result)
```

The p value is less than 0.05, so the difference is statisically significant 

##Overall Candy Ranking

>Q13. What are the five least liked candy types in this set?

```{r}
least_liked <- head(candy[order(candy$winpercent), ], n = 5)
print(least_liked)
```
The 5 least liked candies Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbuster.

>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
top_favorites <- head(candy[order(-candy$winpercent),], n=5)
print(top_favorites)
```
The top 5 favorite candies are Reese's Peanut Butter Cup, Reese's Miniatures, Twix, Kit Kat, Snickers.

>Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
if (!require("ggplot2")) install.packages("ggplot2", dependencies = TRUE)
library(ggplot2)
```


```{r}
ggplot(candy, aes(x = winpercent, y = reorder(rownames(candy), winpercent))) +
  geom_bar(stat = "identity") +
  labs(title = "Candy Ranking by Winpercent", 
       x = "Winpercent", 
       y = "Candy Types") +
  theme_minimal()
```

>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy, aes(x = winpercent, y = reorder(rownames(candy), winpercent))) +
  geom_bar(stat = "identity") +
  labs(title = "Candy Ranking by Winpercent", 
       x = "Winpercent", 
       y = "Candy Types") +
  theme_minimal() 
```

##Add useful color
```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols)
```

>Q17. What is the worst ranked chocolate candy?

```{r}
worst_chocolate <- candy[candy$chocolate == 1, ]  
worst_chocolate <-worst_chocolate[order(worst_chocolate$winpercent), ]  
worst_chocolate <- head(worst_chocolate, 1)
print(worst_chocolate)
```
The worst chocolate are Sixlets.

>Q18. What is the best ranked fruity candy?

```{r}
best_fruity <- candy[candy$fruity == 1, ]  
best_fruity <-best_fruity[order(-best_fruity$winpercent), ]  
best_fruity <- head(best_fruity, 1)
print(best_fruity)
```
The best fruity candy are Starbursts.

##Taking a look at pricepercent: 

```{r}
library(ggrepel)
```

```{r}
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
candy$bang_for_buck <- candy$winpercent / candy$pricepercent
best_value_candy <- candy[which.max(candy$bang_for_buck), ]
print(best_value_candy)
```
The best candy that offers the most bang for your buck are the Tootsie Roll Midgies. 

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```


```{r}
top_expensive <- candy[order(-candy$pricepercent), ][1:5, ]
least_popular_expensive <- top_expensive[which.min(top_expensive$winpercent), ]
print(least_popular_expensive)
```

The most expensive candies and the least populat among them are the Nik L Nip.


>Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment()

```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```
##Exploring the correlation structure 

```{r}
library(corrplot)
```
```{r}
cij <- cor(candy)
corrplot(cij)
```
>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

```{r}
anti_correlated <- which(cij < -0.5, arr.ind = TRUE) 
anti_correlated_vars <- data.frame(
    Var1 = rownames(cij)[anti_correlated[, 1]],
    Var2 = colnames(cij)[anti_correlated[, 2]],
    Correlation = cij[anti_correlated]
)
print(anti_correlated_vars)
```

The two variables that are anti-correlated are fruity and chocolate or fruity and bar, or bar and pluribus. 

>Q23. Similarly, what two variables are most positively correlated?

```{r}
positive_correlated <- which(cij > 0.5, arr.ind = TRUE)
positive_correlated_vars <- data.frame(
    Var1 = rownames(cij)[positive_correlated[, 1]],
    Var2 = colnames(cij)[positive_correlated[, 2]],
    Correlation = cij[positive_correlated]
)
print(positive_correlated_vars)
```
The variables that are most positively correlated are when Var 1 = Var 2, so the correlation is 1 -- for example, chocolate and chocolate have 1 correlation. 

##PCA

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

```{r}
plot(pca$x[, 1:2], 
     xlab = "PC1", 
     ylab = "PC2")
```



```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3, col=my_cols, max.overlaps = 50)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
library(plotly)
```

```{r}
ggplotly(p)
```


```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

```{r}
pca$rotation[,1]
```
This does make sense with what the chart is showing due to how all the positive numbers related to the fruity, hard, pluribus, and bang for buck categories -- this is evident in both the numerical and bar graph data. These 4 categories are seen to be positive for PC1. 
